---
title: "Ryu EHP 2024 LME analyses with Study Group, COPD, and Age"
author: "Min Hyung Ryu"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    code_folding: hide
    df_print: paged
    toc: true
    toc_float: true
---

# Purpose

Purpose of this html document is to LME analyses performed in Ryu et al EHP 2024 manuscript.
The analysis presented here are focused on testing exposure-by-study group, exposure-by-COPD, or exposure-by-age interactions.

Please consider citing our manuscript when using our dataset. Most up-to-date citation recommendation and data use will be posted on our GitHub page.

Our GitHub repository is <https://github.com/COERDUBC/Ryu_EHP_Letter_2024>

For any issue regarding code and data, please post issues to our GitHub page. For any other inquiry, please contact the corresponding author, Dr. Chris Carlsten.

# Manuscript

**Relationship between airway microbiome and the immune response to diesel exhaust: a randomized crossover controlled exposure study**

Min Hyung Ryu^1,2,3^, Illiassou Hamidou Soumana^1,2^, Denise J Wooding^1,2^, Fernando Sergio Leitao Filho^3^, Julia Yang^3^, Corey Nislow^4^, Christopher F Rider^1,2^, Janice M Leung^3^, Chris Carlsten^1,2,3^

^1^ Air Pollution Exposure Lab, Division of Respiratory Medicine, ^2^ Centre for Lung Health, Vancouver Coastal Health Research Institute, ^3^ Heart and Lung Innovation, St. Paul’s Hospital, ^4^ Department of Biochemistry and Molecular Biology, University of British Columbia, Vancouver, BC, Canada

**Correspondence to be addressed to:**\
\
Chris Carlsten, MD, MPH

2775 Laurel St. 7^th^ Floor, The Lung Center, Vancouver General Hospital – Gordon and Leslie Diamond Health Care Centre, Vancouver, BC V5Z 1M9

E-mail: carlsten\@mail.ubc.ca

Telephone: 1-604-875-4729


**Funding:** This study was supported by a research grant from the Canadian Respiratory Research Network.


```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "..")
getwd()

#load libraries
library(here)
library(readxl)
library(writexl)
library(plyr)
library(tidyverse)
library(gmodels)
library(nlme)
library(rmcorr)
library(gtools)
library(DT)
library(tableone)
library(ggpubr)


# set date as an object
currentDate <- format(Sys.Date(), "%Y%m%d")

# set directories
path = here::here() #this should be the parent directory 
data_path = here::here("data")
code_path = here::here("code")
result_path = here::here("results")

# make directory for results on currentDate
if(dir.exists(paste0(result_path,"/", currentDate, "/")) == TRUE){
  print(paste0(result_path,"/", currentDate, "/"))
} else {
  dir.create(paste0(result_path,"/", currentDate, "/"), recursive = TRUE)
}

result_save <- paste0(result_path,"/", currentDate, "/", currentDate, "_")

```

# Data

```{r}
Demographics <- readRDS(paste0(data_path, "/COPA_EHP_Demographics_Public.rds"))
Microbiome <- readRDS(paste0(data_path, "/COPA_EHP_Microbiome_Public.rds"))
colnames(Microbiome)[3] <- "ASV"
Cytokines <- readRDS(paste0(data_path, "/COPA_EHP_VPlex_BAL_Public.rds"))
LODs <- readRDS(paste0(data_path, "/COPA_EHP_VPlex_BAL_LODs_Public.rds"))
colnames(LODs) <- c("Target","LLOD")
Spiro <- readRDS(paste0(data_path, "/COPA_EHP_Spiro_Public.rds"))
```

## Merge data

Merge demographic and Microbiome and add COPD column
```{r}
## Add in COPD information
Demographics$COPD <- as.character(Demographics$Group)

for (i in 1:length(Demographics$EHP_sid)){
  if(Demographics$COPD[i] %in% c("ES", "NS")){
    Demographics$COPD[i] = c("noCOPD")}
  else{
    Demographics$COPD[i] = c("COPD")}}

Demographics$COPD <- factor(Demographics$COPD, level = c("noCOPD", "COPD"))

Demo_microbiome <- merge(Demographics, Microbiome)
Demographics$Group <- factor(Demographics$Group, levels = c("ES","NS", "ES-COPD"))
```

Process spirometry data
```{r}
## taking delta values from the baseline
Spiro_temp <- Spiro
## write a function for finding delta
SubtractMinusFour = function(data){
  arrange(data, EHP_sid, Exposure, Time)
  tempList = list()
  for (i in 1:length(data$EHP_sid)){
    if (data$Time[i] != c("-4h")){
      newRow = data[i, ]
      
      minusFour = data[which(data$EHP_sid == newRow$EHP_sid, data$Exposure == newRow$Exposure), ]
      minusFour = minusFour[which(minusFour$Time == c("-4h")), ]
      
      for(k in 1:nrow(newRow)){  
        for (j in 6:ncol(newRow)){
          newRow[k,j] = ((newRow[k,j] - minusFour[k,j]))
          tempList[[i]] = newRow
        }}}
    
    if (data$Time[i] == c("-4h")){
      tempList[[i]] = NA
    }}
  rm(minusFour, newRow)
  
  newDataFrame = do.call(rbind.data.frame, tempList)
  return(newDataFrame)
}

Spiro_delta = SubtractMinusFour(Spiro_temp)
Spiro_delta = Spiro_delta[-which(is.na(Spiro_delta)), ]
rownames(Spiro_delta) = NULL

Spiro_delta <- merge(Demo_microbiome, Spiro_delta)

Spiro0h = filter(Spiro_delta, Time == "0h")
Spiro1.5h = filter(Spiro_delta, Time == "1.5h")
Spiro24h = filter(Spiro_delta, Time == "24h")

```

Merge cytokine data with demographics and microbiome data
```{r}
## Create cytokine dataframe
Cytokine_data <- merge(Demo_microbiome, Cytokines)
Cytokine_data$Time <- "24h"
```

# Spirometry Analysis

## Study Group
We recruited three groups of participants: never-smokers without COPD (NS group), ex-smokers (minimum 10 pack-years of smoking history) without COPD (ES group), and ex-smokers with mild or moderate COPD (FEV1/FVC < 0.7 and FEV1 % predicted (FEV1pp) >50) (ES-COPD group).

Testing the exposure-by-Group interaction across the entire participants for FEV1. The reference group is the ES group for the following analysis. 

### LME analyses for FEV1 at 0h
```{r}
temp.var.names = c("FEV1 %Pred")
Spiro_data <- Spiro0h
#Demographics$Group <- factor(Demographics$Group, levels = c("ES","NS", "ES-COPD"))

count = 0
for(i in c(1:length(temp.var.names))){
  #i = 1
  tmp.data = Spiro_data[, c(temp.var.names[i], "EHP_sid", "Exposure", "Time", "Group","COPD")]
  tmp.data$Value = tmp.data[,temp.var.names[i]]
  for(j in unique(tmp.data$Time)){
    # j = "24h"
    tmp.data$Exposure = factor(tmp.data$Exposure, levels = c( "FA", "DE"))
    tmp.data$Group <- factor(tmp.data$Group, levels = c("ES","NS", "ES-COPD"))
    tmp.data1 = tmp.data[tmp.data$Time == j,]
    fm1 <- lme(Value ~ Exposure, 
             data = tmp.data1, random = ~1|EHP_sid , method = 'REML', na.action = na.omit)
  count = count + 1
    temp.result = data.frame(ci(fm1))#summary(fm1)$tTable)
  temp.result$Contrast = row.names(temp.result)
  #temp.result = temp.result[!temp.result$Contrast=="(Intercept)", ] 
  temp.result = temp.result[grepl("Exposure", temp.result$Contrast), ]   
  #temp.result$Contrast[temp.result$Contrast=="(Intercept)"]= "FA-S"
  temp.result$Contrast = gsub("Exposure", "", temp.result$Contrast)
  temp.result$Contrast[!temp.result$Contrast =="FA"] =
    paste(temp.result$Contrast[!temp.result$Contrast =="FA"],
          "vs. FA")
  temp.result$Var = temp.var.names[i]
  temp.result$Time = j
  temp.result = temp.result[,c("Time", "Var", "Contrast", "Estimate", "Std..Error", 
                               "DF", "CI.lower", "CI.upper", "p.value")]
    
  if(count == 1){summary.result = temp.result} else {
    summary.result = rbind(summary.result, temp.result)}
  
  fm2 <- lme(Value ~ Exposure*Group , 
             data = tmp.data1, random = ~1 |EHP_sid, method = 'REML', na.action = na.omit)
  
  temp.result = data.frame(ci(fm2))#summary(fm2)$tTable)
  temp.result$Contrast = row.names(temp.result)
  temp.result = temp.result[grepl("ExposureDE:Group", temp.result$Contrast), ] 
  
  temp.result$Contrast = gsub("ExposureDE:Group", "", temp.result$Contrast)
  temp.result$Contrast = paste(temp.result$Contrast, "interaction")
  
  temp.result$Var = temp.var.names[i]
  temp.result$Time = j
  temp.result = temp.result[,c("Time", "Var", "Contrast", "Estimate", "Std..Error", 
                               "DF", "CI.lower", "CI.upper", "p.value")]
  summary.result = rbind(summary.result, temp.result)}
} # end of for i

row.names(summary.result) = c(1:dim(summary.result)[1]) 

## add in the columns with rounding of numbers for nice presentation
summary.result$"Effects [95% CI]" = paste0(round(summary.result$Estimate,2), " [", round(summary.result$CI.lower,2), " to ",  round(summary.result$CI.upper,2),"]")
summary.result$"P value" <- round(summary.result$p.value,4)
summary.result <- summary.result[,-c(4,5,7,8,9)]

summary.result %>% 
  datatable()

```

### LME analyses for FEV1 at 1.5h
```{r}
temp.var.names = c("FEV1 %Pred")
Spiro_data <- Spiro1.5h
#Demographics$Group <- factor(Demographics$Group, levels = c("ES","NS", "ES-COPD"))

count = 0
for(i in c(1:length(temp.var.names))){
  #i = 1
  tmp.data = Spiro_data[, c(temp.var.names[i], "EHP_sid", "Exposure", "Time", "Group","COPD")]
  tmp.data$Value = tmp.data[,temp.var.names[i]]
  for(j in unique(tmp.data$Time)){
    # j = "24h"
    tmp.data$Exposure = factor(tmp.data$Exposure, levels = c( "FA", "DE"))
    tmp.data$Group <- factor(tmp.data$Group, levels = c("ES","NS", "ES-COPD"))
    tmp.data1 = tmp.data[tmp.data$Time == j,]
    fm1 <- lme(Value ~ Exposure, 
             data = tmp.data1, random = ~1|EHP_sid , method = 'REML', na.action = na.omit)
  count = count + 1
    temp.result = data.frame(ci(fm1))#summary(fm1)$tTable)
  temp.result$Contrast = row.names(temp.result)
  #temp.result = temp.result[!temp.result$Contrast=="(Intercept)", ] 
  temp.result = temp.result[grepl("Exposure", temp.result$Contrast), ]   
  #temp.result$Contrast[temp.result$Contrast=="(Intercept)"]= "FA-S"
  temp.result$Contrast = gsub("Exposure", "", temp.result$Contrast)
  temp.result$Contrast[!temp.result$Contrast =="FA"] =
    paste(temp.result$Contrast[!temp.result$Contrast =="FA"],
          "vs. FA")
  temp.result$Var = temp.var.names[i]
  temp.result$Time = j
  temp.result = temp.result[,c("Time", "Var", "Contrast", "Estimate", "Std..Error", 
                               "DF", "CI.lower", "CI.upper", "p.value")]
    
  if(count == 1){summary.result = temp.result} else {
    summary.result = rbind(summary.result, temp.result)}
  
  fm2 <- lme(Value ~ Exposure*Group , 
             data = tmp.data1, random = ~1 |EHP_sid, method = 'REML', na.action = na.omit)
  
  temp.result = data.frame(ci(fm2))#summary(fm2)$tTable)
  temp.result$Contrast = row.names(temp.result)
  temp.result = temp.result[grepl("ExposureDE:Group", temp.result$Contrast), ] 
  
  temp.result$Contrast = gsub("ExposureDE:Group", "", temp.result$Contrast)
  temp.result$Contrast = paste(temp.result$Contrast, "interaction")
  
  temp.result$Var = temp.var.names[i]
  temp.result$Time = j
  temp.result = temp.result[,c("Time", "Var", "Contrast", "Estimate", "Std..Error", 
                               "DF", "CI.lower", "CI.upper", "p.value")]
  summary.result = rbind(summary.result, temp.result)}
} # end of for i

row.names(summary.result) = c(1:dim(summary.result)[1]) 

## add in the columns with rounding of numbers for nice presentation
summary.result$"Effects [95% CI]" = paste0(round(summary.result$Estimate,2), " [", round(summary.result$CI.lower,2), " to ",  round(summary.result$CI.upper,2),"]")
summary.result$"P value" <- round(summary.result$p.value,4)
summary.result <- summary.result[,-c(4,5,7,8,9)]

summary.result %>% 
  datatable()

```

### LME analyses for FEV1 at 24h
```{r}
temp.var.names = c("FEV1 %Pred")
Spiro_data <- Spiro24h
#Demographics$Group <- factor(Demographics$Group, levels = c("ES","NS", "ES-COPD"))

count = 0
for(i in c(1:length(temp.var.names))){
  #i = 1
  tmp.data = Spiro_data[, c(temp.var.names[i], "EHP_sid", "Exposure", "Time", "Group","COPD")]
  tmp.data$Value = tmp.data[,temp.var.names[i]]
  for(j in unique(tmp.data$Time)){
    # j = "24h"
    tmp.data$Exposure = factor(tmp.data$Exposure, levels = c( "FA", "DE"))
    tmp.data$Group <- factor(tmp.data$Group, levels = c("ES","NS", "ES-COPD"))
    tmp.data1 = tmp.data[tmp.data$Time == j,]
    fm1 <- lme(Value ~ Exposure, 
             data = tmp.data1, random = ~1|EHP_sid , method = 'REML', na.action = na.omit)
  count = count + 1
    temp.result = data.frame(ci(fm1))#summary(fm1)$tTable)
  temp.result$Contrast = row.names(temp.result)
  #temp.result = temp.result[!temp.result$Contrast=="(Intercept)", ] 
  temp.result = temp.result[grepl("Exposure", temp.result$Contrast), ]   
  #temp.result$Contrast[temp.result$Contrast=="(Intercept)"]= "FA-S"
  temp.result$Contrast = gsub("Exposure", "", temp.result$Contrast)
  temp.result$Contrast[!temp.result$Contrast =="FA"] =
    paste(temp.result$Contrast[!temp.result$Contrast =="FA"],
          "vs. FA")
  temp.result$Var = temp.var.names[i]
  temp.result$Time = j
  temp.result = temp.result[,c("Time", "Var", "Contrast", "Estimate", "Std..Error", 
                               "DF", "CI.lower", "CI.upper", "p.value")]
    
  if(count == 1){summary.result = temp.result} else {
    summary.result = rbind(summary.result, temp.result)}
  
  fm2 <- lme(Value ~ Exposure*Group , 
             data = tmp.data1, random = ~1 |EHP_sid, method = 'REML', na.action = na.omit)
  
  temp.result = data.frame(ci(fm2))#summary(fm2)$tTable)
  temp.result$Contrast = row.names(temp.result)
  temp.result = temp.result[grepl("ExposureDE:Group", temp.result$Contrast), ] 
  
  temp.result$Contrast = gsub("ExposureDE:Group", "", temp.result$Contrast)
  temp.result$Contrast = paste(temp.result$Contrast, "interaction")
  
  temp.result$Var = temp.var.names[i]
  temp.result$Time = j
  temp.result = temp.result[,c("Time", "Var", "Contrast", "Estimate", "Std..Error", 
                               "DF", "CI.lower", "CI.upper", "p.value")]
  summary.result = rbind(summary.result, temp.result)}
} # end of for i

row.names(summary.result) = c(1:dim(summary.result)[1]) 

## add in the columns with rounding of numbers for nice presentation
summary.result$"Effects [95% CI]" = paste0(round(summary.result$Estimate,2), " [", round(summary.result$CI.lower,2), " to ",  round(summary.result$CI.upper,2),"]")
summary.result$"P value" <- round(summary.result$p.value,4)
summary.result <- summary.result[,-c(4,5,7,8,9)]

summary.result %>% 
  datatable()

```

## COPD status
Testing the exposure-by-COPD interaction across the entire participants for FEV1.

### LME analyses for FEV1 at 0h
```{r}
temp.var.names = c("FEV1 %Pred")
Spiro_data <- Spiro0h

count = 0
for(i in c(1:length(temp.var.names))){
  #i = 1
  tmp.data = Spiro_data[, c(temp.var.names[i], "EHP_sid", "Exposure", "Time", "COPD")]
  tmp.data$Value = tmp.data[,temp.var.names[i]]
  for(j in unique(tmp.data$Time)){
    # j = "24h"
    tmp.data$Exposure = factor(tmp.data$Exposure, levels = c( "FA", "DE"))
    tmp.data$COPD <- factor(tmp.data$COPD, levels = c("noCOPD", "COPD"))
    tmp.data1 = tmp.data[tmp.data$Time == j,]
    fm1 <- lme(Value ~ Exposure, 
             data = tmp.data1, random = ~1|EHP_sid , method = 'REML', na.action = na.omit)
  count = count + 1
    temp.result = data.frame(ci(fm1))#summary(fm1)$tTable)
  temp.result$Contrast = row.names(temp.result)
  #temp.result = temp.result[!temp.result$Contrast=="(Intercept)", ] 
  temp.result = temp.result[grepl("Exposure", temp.result$Contrast), ]   
  #temp.result$Contrast[temp.result$Contrast=="(Intercept)"]= "FA-S"
  temp.result$Contrast = gsub("Exposure", "", temp.result$Contrast)
  temp.result$Contrast[!temp.result$Contrast =="FA"] =
    paste(temp.result$Contrast[!temp.result$Contrast =="FA"],
          "vs. FA")
  temp.result$Var = temp.var.names[i]
  temp.result$Time = j
  temp.result = temp.result[,c("Time", "Var", "Contrast", "Estimate", "Std..Error", 
                               "DF", "CI.lower", "CI.upper", "p.value")]
    
  if(count == 1){summary.result = temp.result} else {
    summary.result = rbind(summary.result, temp.result)}
  
  fm2 <- lme(Value ~ Exposure*COPD , 
             data = tmp.data1, random = ~1 |EHP_sid, method = 'REML', na.action = na.omit)
  
  temp.result = data.frame(ci(fm2))#summary(fm2)$tTable)
  temp.result$Contrast = row.names(temp.result)
  temp.result = temp.result[grepl("ExposureDE:COPD", temp.result$Contrast), ] 
  
  temp.result$Contrast = gsub("ExposureDE:COPD", "", temp.result$Contrast)
  temp.result$Contrast = paste(temp.result$Contrast, "interaction")
  
  temp.result$Var = temp.var.names[i]
  temp.result$Time = j
  temp.result = temp.result[,c("Time", "Var", "Contrast", "Estimate", "Std..Error", 
                               "DF", "CI.lower", "CI.upper", "p.value")]
  summary.result = rbind(summary.result, temp.result)}
} # end of for i

row.names(summary.result) = c(1:dim(summary.result)[1]) 

## add in the columns with rounding of numbers for nice presentation
summary.result$"Effects [95% CI]" = paste0(round(summary.result$Estimate,2), " [", round(summary.result$CI.lower,2), " to ",  round(summary.result$CI.upper,2),"]")
summary.result$"P value" <- round(summary.result$p.value,4)
summary.result <- summary.result[,-c(4,5,7,8,9)]

summary.result %>% 
  datatable()

```

### LME analyses for FEV1 at 1.5h
```{r}
temp.var.names = c("FEV1 %Pred")
Spiro_data <- Spiro1.5h

count = 0
for(i in c(1:length(temp.var.names))){
  #i = 1
  tmp.data = Spiro_data[, c(temp.var.names[i], "EHP_sid", "Exposure", "Time", "COPD")]
  tmp.data$Value = tmp.data[,temp.var.names[i]]
  for(j in unique(tmp.data$Time)){
    # j = "24h"
    tmp.data$Exposure = factor(tmp.data$Exposure, levels = c( "FA", "DE"))
    tmp.data$COPD <- factor(tmp.data$COPD, levels = c("noCOPD", "COPD"))
    tmp.data1 = tmp.data[tmp.data$Time == j,]
    fm1 <- lme(Value ~ Exposure, 
             data = tmp.data1, random = ~1|EHP_sid , method = 'REML', na.action = na.omit)
  count = count + 1
    temp.result = data.frame(ci(fm1))#summary(fm1)$tTable)
  temp.result$Contrast = row.names(temp.result)
  #temp.result = temp.result[!temp.result$Contrast=="(Intercept)", ] 
  temp.result = temp.result[grepl("Exposure", temp.result$Contrast), ]   
  #temp.result$Contrast[temp.result$Contrast=="(Intercept)"]= "FA-S"
  temp.result$Contrast = gsub("Exposure", "", temp.result$Contrast)
  temp.result$Contrast[!temp.result$Contrast =="FA"] =
    paste(temp.result$Contrast[!temp.result$Contrast =="FA"],
          "vs. FA")
  temp.result$Var = temp.var.names[i]
  temp.result$Time = j
  temp.result = temp.result[,c("Time", "Var", "Contrast", "Estimate", "Std..Error", 
                               "DF", "CI.lower", "CI.upper", "p.value")]
    
  if(count == 1){summary.result = temp.result} else {
    summary.result = rbind(summary.result, temp.result)}
  
  fm2 <- lme(Value ~ Exposure*COPD , 
             data = tmp.data1, random = ~1 |EHP_sid, method = 'REML', na.action = na.omit)
  
  temp.result = data.frame(ci(fm2))#summary(fm2)$tTable)
  temp.result$Contrast = row.names(temp.result)
  temp.result = temp.result[grepl("ExposureDE:COPD", temp.result$Contrast), ] 
  
  temp.result$Contrast = gsub("ExposureDE:COPD", "", temp.result$Contrast)
  temp.result$Contrast = paste(temp.result$Contrast, "interaction")
  
  temp.result$Var = temp.var.names[i]
  temp.result$Time = j
  temp.result = temp.result[,c("Time", "Var", "Contrast", "Estimate", "Std..Error", 
                               "DF", "CI.lower", "CI.upper", "p.value")]
  summary.result = rbind(summary.result, temp.result)}
} # end of for i

row.names(summary.result) = c(1:dim(summary.result)[1]) 

## add in the columns with rounding of numbers for nice presentation
summary.result$"Effects [95% CI]" = paste0(round(summary.result$Estimate,2), " [", round(summary.result$CI.lower,2), " to ",  round(summary.result$CI.upper,2),"]")
summary.result$"P value" <- round(summary.result$p.value,4)
summary.result <- summary.result[,-c(4,5,7,8,9)]

summary.result %>% 
  datatable()

```

### LME analyses for FEV1 at 24h
```{r}
temp.var.names = c("FEV1 %Pred")
Spiro_data <- Spiro24h

count = 0
for(i in c(1:length(temp.var.names))){
  #i = 1
  tmp.data = Spiro_data[, c(temp.var.names[i], "EHP_sid", "Exposure", "Time", "COPD")]
  tmp.data$Value = tmp.data[,temp.var.names[i]]
  for(j in unique(tmp.data$Time)){
    # j = "24h"
    tmp.data$Exposure = factor(tmp.data$Exposure, levels = c( "FA", "DE"))
    tmp.data$COPD <- factor(tmp.data$COPD, levels = c("noCOPD", "COPD"))
    tmp.data1 = tmp.data[tmp.data$Time == j,]
    fm1 <- lme(Value ~ Exposure, 
             data = tmp.data1, random = ~1|EHP_sid , method = 'REML', na.action = na.omit)
  count = count + 1
    temp.result = data.frame(ci(fm1))#summary(fm1)$tTable)
  temp.result$Contrast = row.names(temp.result)
  #temp.result = temp.result[!temp.result$Contrast=="(Intercept)", ] 
  temp.result = temp.result[grepl("Exposure", temp.result$Contrast), ]   
  #temp.result$Contrast[temp.result$Contrast=="(Intercept)"]= "FA-S"
  temp.result$Contrast = gsub("Exposure", "", temp.result$Contrast)
  temp.result$Contrast[!temp.result$Contrast =="FA"] =
    paste(temp.result$Contrast[!temp.result$Contrast =="FA"],
          "vs. FA")
  temp.result$Var = temp.var.names[i]
  temp.result$Time = j
  temp.result = temp.result[,c("Time", "Var", "Contrast", "Estimate", "Std..Error", 
                               "DF", "CI.lower", "CI.upper", "p.value")]
    
  if(count == 1){summary.result = temp.result} else {
    summary.result = rbind(summary.result, temp.result)}
  
  fm2 <- lme(Value ~ Exposure*COPD , 
             data = tmp.data1, random = ~1 |EHP_sid, method = 'REML', na.action = na.omit)
  
  temp.result = data.frame(ci(fm2))#summary(fm2)$tTable)
  temp.result$Contrast = row.names(temp.result)
  temp.result = temp.result[grepl("ExposureDE:COPD", temp.result$Contrast), ] 
  
  temp.result$Contrast = gsub("ExposureDE:COPD", "", temp.result$Contrast)
  temp.result$Contrast = paste(temp.result$Contrast, "interaction")
  
  temp.result$Var = temp.var.names[i]
  temp.result$Time = j
  temp.result = temp.result[,c("Time", "Var", "Contrast", "Estimate", "Std..Error", 
                               "DF", "CI.lower", "CI.upper", "p.value")]
  summary.result = rbind(summary.result, temp.result)}
} # end of for i

row.names(summary.result) = c(1:dim(summary.result)[1]) 

## add in the columns with rounding of numbers for nice presentation
summary.result$"Effects [95% CI]" = paste0(round(summary.result$Estimate,2), " [", round(summary.result$CI.lower,2), " to ",  round(summary.result$CI.upper,2),"]")
summary.result$"P value" <- round(summary.result$p.value,4)
summary.result <- summary.result[,-c(4,5,7,8,9)]

summary.result %>% 
  datatable()

```

## Age
Testing the exposure-by-age interaction across the entire participants for FEV1.

### LME analyses for FEV1 at 0h
```{r}
temp.var.names = c("FEV1 %Pred")
Spiro_data <- Spiro0h

count = 0
for(i in c(1:length(temp.var.names))){
  #i = 1
  tmp.data = Spiro_data[, c(temp.var.names[i], "EHP_sid", "Exposure", "Time", "Age","COPD")]
  tmp.data$Value = tmp.data[,temp.var.names[i]]
  for(j in unique(tmp.data$Time)){
    # j = "24h"
    tmp.data$Exposure = factor(tmp.data$Exposure, levels = c( "FA", "DE"))
    tmp.data1 = tmp.data[tmp.data$Time == j,]
    fm1 <- lme(Value ~ Exposure, 
             data = tmp.data1, random = ~1|EHP_sid , method = 'REML', na.action = na.omit)
  count = count + 1
    temp.result = data.frame(ci(fm1))#summary(fm1)$tTable)
  temp.result$Contrast = row.names(temp.result)
  #temp.result = temp.result[!temp.result$Contrast=="(Intercept)", ] 
  temp.result = temp.result[grepl("Exposure", temp.result$Contrast), ]   
  #temp.result$Contrast[temp.result$Contrast=="(Intercept)"]= "FA-S"
  temp.result$Contrast = gsub("Exposure", "", temp.result$Contrast)
  temp.result$Contrast[!temp.result$Contrast =="FA"] =
    paste(temp.result$Contrast[!temp.result$Contrast =="FA"],
          "vs. FA")
  temp.result$Var = temp.var.names[i]
  temp.result$Time = j
  temp.result = temp.result[,c("Time", "Var", "Contrast", "Estimate", "Std..Error", 
                               "DF", "CI.lower", "CI.upper", "p.value")]
    
  if(count == 1){summary.result = temp.result} else {
    summary.result = rbind(summary.result, temp.result)}
  
  fm2 <- lme(Value ~ Exposure*Age , 
             data = tmp.data1, random = ~1 |EHP_sid, method = 'REML', na.action = na.omit)
  
  temp.result = data.frame(ci(fm2))#summary(fm2)$tTable)
  temp.result$Contrast = row.names(temp.result)
  temp.result = temp.result[grepl("ExposureDE:Age", temp.result$Contrast), ] 
  
  temp.result$Contrast = gsub("ExposureDE:", "", temp.result$Contrast)
  temp.result$Contrast = paste(temp.result$Contrast, "interaction")
  
  temp.result$Var = temp.var.names[i]
  temp.result$Time = j
  temp.result = temp.result[,c("Time", "Var", "Contrast", "Estimate", "Std..Error", 
                               "DF", "CI.lower", "CI.upper", "p.value")]
  summary.result = rbind(summary.result, temp.result)}
} # end of for i

row.names(summary.result) = c(1:dim(summary.result)[1]) 

## add in the columns with rounding of numbers for nice presentation
summary.result$"Effects [95% CI]" = paste0(round(summary.result$Estimate,2), " [", round(summary.result$CI.lower,2), " to ",  round(summary.result$CI.upper,2),"]")
summary.result$"P value" <- round(summary.result$p.value,4)
summary.result <- summary.result[,-c(4,5,7,8,9)]

summary.result %>% 
  datatable()

```

### LME analyses for FEV1 at 1.5h
```{r}
temp.var.names = c("FEV1 %Pred")
Spiro_data <- Spiro1.5h

count = 0
for(i in c(1:length(temp.var.names))){
  #i = 1
  tmp.data = Spiro_data[, c(temp.var.names[i], "EHP_sid", "Exposure", "Time", "Age","COPD")]
  tmp.data$Value = tmp.data[,temp.var.names[i]]
  for(j in unique(tmp.data$Time)){
    # j = "24h"
    tmp.data$Exposure = factor(tmp.data$Exposure, levels = c( "FA", "DE"))
    tmp.data1 = tmp.data[tmp.data$Time == j,]
    fm1 <- lme(Value ~ Exposure, 
             data = tmp.data1, random = ~1|EHP_sid , method = 'REML', na.action = na.omit)
  count = count + 1
    temp.result = data.frame(ci(fm1))#summary(fm1)$tTable)
  temp.result$Contrast = row.names(temp.result)
  #temp.result = temp.result[!temp.result$Contrast=="(Intercept)", ] 
  temp.result = temp.result[grepl("Exposure", temp.result$Contrast), ]   
  #temp.result$Contrast[temp.result$Contrast=="(Intercept)"]= "FA-S"
  temp.result$Contrast = gsub("Exposure", "", temp.result$Contrast)
  temp.result$Contrast[!temp.result$Contrast =="FA"] =
    paste(temp.result$Contrast[!temp.result$Contrast =="FA"],
          "vs. FA")
  temp.result$Var = temp.var.names[i]
  temp.result$Time = j
  temp.result = temp.result[,c("Time", "Var", "Contrast", "Estimate", "Std..Error", 
                               "DF", "CI.lower", "CI.upper", "p.value")]
    
  if(count == 1){summary.result = temp.result} else {
    summary.result = rbind(summary.result, temp.result)}
  
  fm2 <- lme(Value ~ Exposure*Age , 
             data = tmp.data1, random = ~1 |EHP_sid, method = 'REML', na.action = na.omit)
  
  temp.result = data.frame(ci(fm2))#summary(fm2)$tTable)
  temp.result$Contrast = row.names(temp.result)
  temp.result = temp.result[grepl("ExposureDE:Age", temp.result$Contrast), ] 
  
  temp.result$Contrast = gsub("ExposureDE:", "", temp.result$Contrast)
  temp.result$Contrast = paste(temp.result$Contrast, "interaction")
  
  temp.result$Var = temp.var.names[i]
  temp.result$Time = j
  temp.result = temp.result[,c("Time", "Var", "Contrast", "Estimate", "Std..Error", 
                               "DF", "CI.lower", "CI.upper", "p.value")]
  summary.result = rbind(summary.result, temp.result)}
} # end of for i

row.names(summary.result) = c(1:dim(summary.result)[1]) 

## add in the columns with rounding of numbers for nice presentation
summary.result$"Effects [95% CI]" = paste0(round(summary.result$Estimate,2), " [", round(summary.result$CI.lower,2), " to ",  round(summary.result$CI.upper,2),"]")
summary.result$"P value" <- round(summary.result$p.value,4)
summary.result <- summary.result[,-c(4,5,7,8,9)]

summary.result %>% 
  datatable()

```

### LME analyses for FEV1 at 24h
```{r}
temp.var.names = c("FEV1 %Pred")
Spiro_data <- Spiro24h

count = 0
for(i in c(1:length(temp.var.names))){
  #i = 1
  tmp.data = Spiro_data[, c(temp.var.names[i], "EHP_sid", "Exposure", "Time", "Age","COPD")]
  tmp.data$Value = tmp.data[,temp.var.names[i]]
  for(j in unique(tmp.data$Time)){
    # j = "24h"
    tmp.data$Exposure = factor(tmp.data$Exposure, levels = c( "FA", "DE"))
    tmp.data1 = tmp.data[tmp.data$Time == j,]
    fm1 <- lme(Value ~ Exposure, 
             data = tmp.data1, random = ~1|EHP_sid , method = 'REML', na.action = na.omit)
  count = count + 1
    temp.result = data.frame(ci(fm1))#summary(fm1)$tTable)
  temp.result$Contrast = row.names(temp.result)
  #temp.result = temp.result[!temp.result$Contrast=="(Intercept)", ] 
  temp.result = temp.result[grepl("Exposure", temp.result$Contrast), ]   
  #temp.result$Contrast[temp.result$Contrast=="(Intercept)"]= "FA-S"
  temp.result$Contrast = gsub("Exposure", "", temp.result$Contrast)
  temp.result$Contrast[!temp.result$Contrast =="FA"] =
    paste(temp.result$Contrast[!temp.result$Contrast =="FA"],
          "vs. FA")
  temp.result$Var = temp.var.names[i]
  temp.result$Time = j
  temp.result = temp.result[,c("Time", "Var", "Contrast", "Estimate", "Std..Error", 
                               "DF", "CI.lower", "CI.upper", "p.value")]
    
  if(count == 1){summary.result = temp.result} else {
    summary.result = rbind(summary.result, temp.result)}
  
  fm2 <- lme(Value ~ Exposure*Age , 
             data = tmp.data1, random = ~1 |EHP_sid, method = 'REML', na.action = na.omit)
  
  temp.result = data.frame(ci(fm2))#summary(fm2)$tTable)
  temp.result$Contrast = row.names(temp.result)
  temp.result = temp.result[grepl("ExposureDE:Age", temp.result$Contrast), ] 
  
  temp.result$Contrast = gsub("ExposureDE:", "", temp.result$Contrast)
  temp.result$Contrast = paste(temp.result$Contrast, "interaction")
  
  temp.result$Var = temp.var.names[i]
  temp.result$Time = j
  temp.result = temp.result[,c("Time", "Var", "Contrast", "Estimate", "Std..Error", 
                               "DF", "CI.lower", "CI.upper", "p.value")]
  summary.result = rbind(summary.result, temp.result)}
} # end of for i

row.names(summary.result) = c(1:dim(summary.result)[1]) 

## add in the columns with rounding of numbers for nice presentation
summary.result$"Effects [95% CI]" = paste0(round(summary.result$Estimate,2), " [", round(summary.result$CI.lower,2), " to ",  round(summary.result$CI.upper,2),"]")
summary.result$"P value" <- round(summary.result$p.value,4)
summary.result <- summary.result[,-c(4,5,7,8,9)]

summary.result %>% 
  datatable()

```

# Cytokine Analysis

Here we are testing for all the cytokines detected in the BAL samples.
```{r}
#cytokines to be tested
temp.var.names= c("GM-CSF", "IL-1α", "IL-6", "IL-7", "IL-8", "IL-12/IL-23p40", "IL-15", "IL-16", "IP-10", "MCP-1", "MDC", "MIP-1β", "TARC", "VEGF") 
```

## Study Group
#### Main effect analysis and exposure-by-Group interaction test
```{r}
count = 0
for(i in c(1:length(temp.var.names))){
  #i = 1
  tmp.data = Cytokine_data[, c(temp.var.names[i], "EHP_sid", "Exposure", "Time", "Group", "COPD", "Age")]
  tmp.data$Value = tmp.data[,temp.var.names[i]]
  for(j in unique(tmp.data$Time)){
    # j = "24h"
    tmp.data$Exposure = factor(tmp.data$Exposure, levels = c( "FA", "DE"))
    tmp.data$Group <- factor(tmp.data$Group, levels = c("ES","NS", "ES-COPD"))

    tmp.data1 = tmp.data[tmp.data$Time == j,]
    #tmp.data1 = tmp.data1[complete.cases(tmp.data1),]
    fm1 <- lme(log10(Value) ~ Exposure + Age, 
             data = tmp.data1, random = ~1|EHP_sid , method = 'REML', na.action = na.omit)
  count = count + 1
  temp.result = data.frame(ci(fm1))#summary(fm1)$tTable)
  temp.result$Contrast = row.names(temp.result)
  #temp.result = temp.result[!temp.result$Contrast=="(Intercept)", ] 
  temp.result = temp.result[grepl("Exposure", temp.result$Contrast), ]   
  #temp.result$Contrast[temp.result$Contrast=="(Intercept)"]= "FA-S"
  temp.result$Contrast = gsub("Exposure", "", temp.result$Contrast)
  temp.result$Contrast[!temp.result$Contrast =="FA"] =
    paste(temp.result$Contrast[!temp.result$Contrast =="FA"],
          "vs. FA")
  temp.result$Var = temp.var.names[i]
  temp.result$Time = j
  temp.result = temp.result[,c("Time", "Var", "Contrast", "Estimate", "Std..Error", 
                               "DF", "CI.lower", "CI.upper", "p.value")]
    
  if(count == 1){summary.result = temp.result} else {
    summary.result = rbind(summary.result, temp.result)}
  
  fm2 <- lme(log(Value) ~ Exposure*Group + Age, 
             data = tmp.data1, random = ~1 |EHP_sid, method = 'REML')
  
  temp.result = data.frame(ci(fm2))#summary(fm2)$tTable)
  temp.result$Contrast = row.names(temp.result)
  temp.result = temp.result[grepl("ExposureDE:Group", temp.result$Contrast), ] 
  temp.result$Contrast = gsub("ExposureDE:Group", "", temp.result$Contrast)
  temp.result$Contrast = paste(temp.result$Contrast, "interaction")
  
  temp.result$Var = temp.var.names[i]
  temp.result$Time = j
  temp.result = temp.result[,c("Time", "Var", "Contrast", "Estimate", "Std..Error", 
                               "DF", "CI.lower", "CI.upper", "p.value")]
  summary.result = rbind(summary.result, temp.result)
  } # end of for (j)
} # end of for i

row.names(summary.result) = c(1:dim(summary.result)[1])

## add in the columns with rounding of numbers for nice presentation
summary.result$"Effects [95% CI] back transformed" = paste0(round(10^summary.result$Estimate,2), " [", round(10^summary.result$CI.lower,2), " to ",  round(10^summary.result$CI.upper,2),"]")
summary.result$"P value" <- round(summary.result$p.value,4)
summary.result <- summary.result[,-c(4,5,7,8,9)]
summary.result %>% 
  datatable()
```

## COPD Status
#### Main effect analysis and exposure-by-COPD interaction test
```{r}
count = 0
for(i in c(1:length(temp.var.names))){
  #i = 1
  tmp.data = Cytokine_data[, c(temp.var.names[i], "EHP_sid", "Exposure", "Time", "COPD", "Age")]
  tmp.data$Value = tmp.data[,temp.var.names[i]]
  for(j in unique(tmp.data$Time)){
    # j = "24h"
    tmp.data$Exposure = factor(tmp.data$Exposure, levels = c( "FA", "DE"))
    tmp.data$COPD <- factor(tmp.data$COPD, levels = c("noCOPD", "COPD"))

    tmp.data1 = tmp.data[tmp.data$Time == j,]
    #tmp.data1 = tmp.data1[complete.cases(tmp.data1),]
    fm1 <- lme(log10(Value) ~ Exposure + Age, 
             data = tmp.data1, random = ~1|EHP_sid , method = 'REML', na.action = na.omit)
  count = count + 1
  temp.result = data.frame(ci(fm1))#summary(fm1)$tTable)
  temp.result$Contrast = row.names(temp.result)
  #temp.result = temp.result[!temp.result$Contrast=="(Intercept)", ] 
  temp.result = temp.result[grepl("Exposure", temp.result$Contrast), ]   
  #temp.result$Contrast[temp.result$Contrast=="(Intercept)"]= "FA-S"
  temp.result$Contrast = gsub("Exposure", "", temp.result$Contrast)
  temp.result$Contrast[!temp.result$Contrast =="FA"] =
    paste(temp.result$Contrast[!temp.result$Contrast =="FA"],
          "vs. FA")
  temp.result$Var = temp.var.names[i]
  temp.result$Time = j
  temp.result = temp.result[,c("Time", "Var", "Contrast", "Estimate", "Std..Error", 
                               "DF", "CI.lower", "CI.upper", "p.value")]
    
  if(count == 1){summary.result = temp.result} else {
    summary.result = rbind(summary.result, temp.result)}
  
  fm2 <- lme(log(Value) ~ Exposure*COPD + Age, 
             data = tmp.data1, random = ~1 |EHP_sid, method = 'REML')
  
  temp.result = data.frame(ci(fm2))#summary(fm2)$tTable)
  temp.result$Contrast = row.names(temp.result)
  temp.result = temp.result[grepl("ExposureDE:COPD", temp.result$Contrast), ] 
  temp.result$Contrast = gsub("ExposureDE:COPD", "", temp.result$Contrast)
  temp.result$Contrast = paste(temp.result$Contrast, "interaction")
  
  temp.result$Var = temp.var.names[i]
  temp.result$Time = j
  temp.result = temp.result[,c("Time", "Var", "Contrast", "Estimate", "Std..Error", 
                               "DF", "CI.lower", "CI.upper", "p.value")]
  summary.result = rbind(summary.result, temp.result)
  } # end of for (j)
} # end of for i

row.names(summary.result) = c(1:dim(summary.result)[1])

## add in the columns with rounding of numbers for nice presentation
summary.result$"Effects [95% CI] back transformed" = paste0(round(10^summary.result$Estimate,2), " [", round(10^summary.result$CI.lower,2), " to ",  round(10^summary.result$CI.upper,2),"]")
summary.result$"P value" <- round(summary.result$p.value,4)
summary.result <- summary.result[,-c(4,5,7,8,9)]
summary.result %>% 
  datatable()
```

## Age
#### Main effect analysis and exposure-by-Age interaction test
```{r}
count = 0
for(i in c(1:length(temp.var.names))){
  #i = 1
  tmp.data = Cytokine_data[, c(temp.var.names[i], "EHP_sid", "Exposure", "Time", "Age")]
  tmp.data$Value = tmp.data[,temp.var.names[i]]
  for(j in unique(tmp.data$Time)){
    # j = "24h"
    tmp.data$Exposure = factor(tmp.data$Exposure, levels = c( "FA", "DE"))

    tmp.data1 = tmp.data[tmp.data$Time == j,]
    #tmp.data1 = tmp.data1[complete.cases(tmp.data1),]
    fm1 <- lme(log10(Value) ~ Exposure + Age, 
             data = tmp.data1, random = ~1|EHP_sid , method = 'REML', na.action = na.omit)
  count = count + 1
  temp.result = data.frame(ci(fm1))#summary(fm1)$tTable)
  temp.result$Contrast = row.names(temp.result)
  #temp.result = temp.result[!temp.result$Contrast=="(Intercept)", ] 
  temp.result = temp.result[grepl("Exposure", temp.result$Contrast), ]   
  #temp.result$Contrast[temp.result$Contrast=="(Intercept)"]= "FA-S"
  temp.result$Contrast = gsub("Exposure", "", temp.result$Contrast)
  temp.result$Contrast[!temp.result$Contrast =="FA"] =
    paste(temp.result$Contrast[!temp.result$Contrast =="FA"],
          "vs. FA")
  temp.result$Var = temp.var.names[i]
  temp.result$Time = j
  temp.result = temp.result[,c("Time", "Var", "Contrast", "Estimate", "Std..Error", 
                               "DF", "CI.lower", "CI.upper", "p.value")]
    
  if(count == 1){summary.result = temp.result} else {
    summary.result = rbind(summary.result, temp.result)}
  
  fm2 <- lme(log(Value) ~ Exposure*Age, 
             data = tmp.data1, random = ~1 |EHP_sid, method = 'REML')
  
  temp.result = data.frame(ci(fm2))#summary(fm2)$tTable)
  temp.result$Contrast = row.names(temp.result)
  temp.result = temp.result[grepl("ExposureDE:Age", temp.result$Contrast), ] 
  temp.result$Contrast = gsub("ExposureDE:", "", temp.result$Contrast)
  temp.result$Contrast = paste(temp.result$Contrast, "interaction")
  
  temp.result$Var = temp.var.names[i]
  temp.result$Time = j
  temp.result = temp.result[,c("Time", "Var", "Contrast", "Estimate", "Std..Error", 
                               "DF", "CI.lower", "CI.upper", "p.value")]
  summary.result = rbind(summary.result, temp.result)
  } # end of for (j)
} # end of for i

row.names(summary.result) = c(1:dim(summary.result)[1])

## add in the columns with rounding of numbers for nice presentation
summary.result$"Effects [95% CI] back transformed" = paste0(round(10^summary.result$Estimate,2), " [", round(10^summary.result$CI.lower,2), " to ",  round(10^summary.result$CI.upper,2),"]")
summary.result$"P value" <- round(summary.result$p.value,4)
summary.result <- summary.result[,-c(4,5,7,8,9)]
summary.result %>% 
  datatable()
```


# Session Info

```{r session information calling, echo=FALSE}
sessionInfo()
```
